---
title: "Back-calculation of R(t)"
author: Tenglong Li, PhD^[Boston University, Biostatistics]; Laura White, PhD^[Boston University, Biostatistics]; Chad Milando, PhD*^[Boston University, Environmental Health, cmilando@bu.edu]
date: "2024-01-03"
output: html_document
runtime: shiny
---

```{=html}
<style type="text/css">
.main-container {
  max-width: 750px;
  margin-left: auto;
  margin-right: auto;
}
</style>
```

```{css, echo=FALSE}
.plotlysize {
  height: 220px;
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
library(purrr)
library(shiny)
library(splines)
library(tidyverse)
```

This R Markdown document walks through the steps of using Bayesian inference to create more realistic estimates of time-dependent reproductive number, $R(t)$ that can be helpful for surveillance and intervention planning of infectious diseases, like COVID-19.

This document has several sections:

-   Simulating line list data
-   Back-calculation and now-casting

### Background
**Line-list data** means you have a single row for each case, that has dates for: infection, symptom onset, positive test, and when this was reported to public health agencies.

Sources for describing these windows:

* 

## Simulating line-list data

Assuming you have 100 initial cases (infectors), and distributions for infections and onset are as described for COVID-19 in the supplemental of [Li and White](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1009210), set the initial $R_0$, the asymptotic $R(t)$, the parameters for the truncated Negative Binomial (NB) distribution that defines the maximum reporting delay distribution, and the modeling window. Perfect instantaneous reporting can be achieved with either a max. reporting delay of 0 days or with a NB mean of 0. NB dispersion (also called 'size') is a measure of over-dispersion (smaller size means more over-dispersion, which means variance != mean). For more information see [Zeileis](https://edoc.unibas.ch/15400/).

```{r get_curves, echo=FALSE}

n0 = 100; 

## INPUTS
inputPanel(
  # row 1
  sliderInput("r0", label = "Initial r0:",
              min = 0.2, max = 2, value = 1, step = 0.1),
  sliderInput("NB_m", label = "NB mean (days)", value = 9, min = 0., max = 20),
  sliderInput("maxdelay", label = "Max. reporting delay (days)", value = 20,
              min = 0, max = 30, step = 1),
  # row2
  sliderInput("RTINF", label = "Final R(t):",
              min = 0.2, max = 2, value = 1, step = 0.1),
  sliderInput("NB_r", label = "NB dispersion", value = 3, min = 0.1, max = 20),
  sliderInput("ld", label = "Modeling window (days)", value = 30,
              min = 0, max = 60, step = 5)
)
```

Using this information, describe the daily infections, onsets, and reporting days of first generation (the first round of infectees):
```{r get_curves2, echo=FALSE}
# set.seed(4)
# wild ** set.seed(1) doesn't work....
# and in general, setting the seed makes this look weird ...
# *****

plot_dat <- reactive({
  # ----------
  # one draw from poisson given by lambda = initial cases * initial reproductive number
  # this gives the first timestep of infectees from the initial cases (?)
  n1 <- rpois(1, n0 * input$r0) 
  # hist(rpois(1000, n0 * r0))
  # n1
  
  # LiT: Date of infection
  # Generation interval distribution draw for n1 infectees
  # generation interval is the time between infection and transmission
  # shape = 4.29 and rate = 1.18 are based on COVID-19 literature
  # > ceiling() rounds up to integer values for days 
  # NOTE: this will be replaced by "serial interval" data later (?)
  inf1 <- rgamma(n1, shape = 4.29, rate = 1.18) 
  i1   <- ceiling(inf1) 
  # hist(i1)
  # length(i1)
  
  # LiT: Date of onset sympotom
  # Incubation period: period between exposure and appearance of symptoms
  # lognormal distribution with mean of 1.621 and sd of 0.418, based on COVID-19 literature
  # > ceiling() rounds up to integer values for days 
  inc1 <- exp(rnorm(n1, 1.621, 0.418))
  o1   <- i1 + ceiling(inc1) 
  # hist(o1)
  # length(o1)

  # LiT: Date of report
  # negative binomial distribution for the reporting delay between
  # testing positive and reporting to public health agencies
  # with size parameter = r and mean(mu) = m and
  # with an upper bound of l (aka `maxdelay`), the max reporting delay
  # there are n = maxdelay + 1 of these values
  p <- dnbinom(0:input$maxdelay, size = input$NB_r, mu = input$NB_m)
  d1 <- o1 + sample(0:input$maxdelay, size = n1, prob = p, replace = TRUE) 
  # hist(d1)
  # length(d1)
  
  # create a combined dataset
  # CM: add generation #
  gen_i <- 1
  gen_i_c <- rep(gen_i, length(i1))
  dat <- cbind(i1, o1, d1, gen_i_c)
  
  dat
})
# 
# # Curves
# # area under the curve should be = 196
renderPlot({
  as_tibble(plot_dat()) %>%
    rename("Daily Infections" = i1,
           "Daily Onsets" = o1,
           "Daily Reports" = d1) %>%
    pivot_longer(cols = c("Daily Infections", "Daily Onsets",
                          "Daily Reports")) %>%
    group_by(name, gen_i_c, value) %>% tally() %>%
    ggplot(.) +
    geom_bar(aes(x = value, fill = factor(gen_i_c), y = n),
             position = 'stack', stat = 'identity', color = 'black',
             lwd = 0.1, fill = '#6495ED') +
    facet_wrap(~name, nrow = 1) +
    ylab('N') + xlab('Day') +
    scale_fill_brewer(name = 'Gen.#', type = 'seq',
                      direction = 1, palette = 3) +
    theme_gray() + coord_cartesian(xlim = c(0, as.numeric(input$ld)),
                                   ylim = c(0, 50)) +
    theme(
      axis.text = element_text(size = 10),
          axis.title = element_text(size = 14),
          strip.text = element_text(size = 14)
    )
}, height = 200)
```

Now, create your $R(t)$ curve so you can simulate the next generation(s) of cases (starting at $t=1$. Pre-defining the shape of $R(t)$ allows you to simulate infectious disease dynamics, interventions, etc. Note: if $R(t) > 1$ for all $t$, the cases will expontentially increase, and the code will not complete for modeling windows greater than 30 days.

<div class = 'plotlysize'>
```{r set_rt, echo=FALSE}
# PLOT PARAMETERS
YMAX <- 2
NPTS <- 6
BSDEG <- 1
XSTART <- 1

# REFERENCE: # https://stackoverflow.com/questions/47280032/draggable-line-chart-in-r-shiny
### asb should be '30'
rv <- reactiveValues(
  x = NULL,
  y = rep(1, length.out = NPTS)
)

observeEvent(input$ld, {
  rv$x <- seq(XSTART, input$ld, length.out = NPTS)
})


rt_grid <- reactive({
  this.x <- seq(min(rv$x), max(rv$x), length = input$ld)
  suppressWarnings(this.y <- predict(this.model(), data.frame(x = this.x)))
  data.frame(x = this.x, y = this.y)
})

this.model <- reactive({
  d <- data.frame(x = rv$x, y = rv$y)
  lm(y ~ bs(x, degree = BSDEG, knots = rv$x), d)
})

# Render plot
renderPlotly({
  
  req(rv$x)
  req(rv$y)
  req(rt_grid()$x)
  req(rt_grid()$y)
  
  # creates a list of circle shapes from x/y data
  circles <- map2(
    rv$x, rv$y,
    ~ list(
      type = "circle",
      # anchor circles
      xanchor = .x,
      yanchor = .y,
      # give each circle a 2 pixel diameter
      x0 = -4, x1 = 4,
      y0 = -4, y1 = 4,
      xsizemode = "pixel",
      ysizemode = "pixel",
      # other visual properties
      fillcolor = "red",
      ## hover info?
      ## stroke??
      line = list(color = "black", stroke = 5)
    )
  )

  # plot the shapes and fitted line
  p <- plot_ly(source="mysource", height=200) %>%
    add_lines(x = c(XSTART, input$ld), y = c(1, 1), color = I("black"),
              line = list(widthh=0.5, dash="dot"), showlegend = F) %>%
    add_lines(x = rt_grid()$x, y = rt_grid()$y, color = I("#F89880"),
              showlegend = F,
              hoverinfo = 'y',
              hovertemplate = paste("%{y:.3f}<extra></extra>")
              ) %>% 
    layout(
        plot_bgcolor = '#f0f0f0', # Light grey background
        # xaxis
        xaxis = list(
          title = "t",
          zerolinecolor = "#f0f0f0", # Light grey zeroline color
          gridcolor = "#ffffff",     # White grid lines
          titlefont = list(family = "Arial, sans-serif"),
          tickfont = list(family = "Arial, sans-serif")
        ),
        # yaxis
        yaxis = list(
          title = "R(t)",
          range = c(0, YMAX),
          zerolinecolor = "#f0f0f0", # Light grey zeroline color
          gridcolor = "#ffffff",     # White grid lines
          titlefont = list(family = "Arial, sans-serif"),
          tickfont = list(family = "Arial, sans-serif")
        ),
        margin = list(l = 35, r = 0, b = 0, t = 0), # Adjusted margins
        shapes = circles,
        font = list(family = "Arial, sans-serif")
      ) %>%
      config(edits = list(shapePosition = TRUE))
  
    event_register(p, "plotly_relayout")
    
    p
})

# update x/y reactive values in response to changes in shape anchors

observe({
  ed <- event_data("plotly_relayout", source = 'mysource')
  shape_anchors <- ed[grepl("^shapes.*anchor$", names(ed))]
  if (length(shape_anchors) != 2) {
    return()
  }
  row_index <- unique(readr::parse_number(names(shape_anchors)) + 1)
  pts <- as.numeric(shape_anchors)
  
  # some controls on the bounds
  if(row_index == 1) {
    rv$x[row_index] <- XSTART
  } else if (row_index == NPTS) {
    rv$x[row_index] <- input$ld
  } else {
    rv$x[row_index] <- sapply(pts[1], function(x) 
      ifelse(x < XSTART, XSTART, ifelse(x > input$ld, input$ld, x)))
  }
  
  rv$y[row_index] <- sapply(pts[2], function(y) 
    ifelse(y < 0, 0, ifelse(y > YMAX, YMAX, y)))
})
```
</div>

Now, create the all subsequent generations within the modeling window (generation # is indicated by fill color in the graph below, with lighter being later generations. The first generation is outlined in red):
```{r echo=FALSE}

dat_all <- reactive({
  
  dat <- as_tibble(plot_dat())

  i1 <- dat$i1
  gen_i <- 1
  p <- dnbinom(0:input$maxdelay, size = input$NB_r, mu = input$NB_m)
  
  # ----------
  # while there are any initial infection dates lower than the modeling window
  # end, simulate the next generation
  #while (gen_i < 2) {
  while(any(i1<=input$ld)) {
    ## new generation
    gen_i <- gen_i + 1

    
    ## Infection dates of previous generation
    date <- as.numeric(names(table(i1)))
    
    ## Number of cases on each date
    no   <- as.numeric(table(i1))
    #no
    # stopifnot(sum(no) == n1)
    ## get the next iteration of rT using the file provided.
    #mu   <- no * rt[date]
    rt = rt_grid()$y
    rt <- c(rt, rep(input$RTINF, max(date))) ## repeat RTINF until the end of time
    mu   <- no * rt[date] # ****** <<<< TODO:
    #mu <- no * r0
    #mu
    #sum(mu)
    
    ## simulate the next generation of infectees
    #n1   <- rpois(1, mu)
    #n1
    n1   <- sapply(mu, function(mx) rpois(1, mx))
    #n1
    #**** ?
    # this doesn't make several n1, it just does the first one....
    # incorrect ??
    
    ## and their generation interval
    inf1 <- rgamma(sum(n1), shape = 4.29, rate = 1.18)
    # include the start dates before the generation time
    i1   <- rep(date, times = n1) + ceiling(inf1)
    ## 
    inc1 <- exp(rnorm(sum(n1), 1.621, 0.418))
    o1   <- i1 + ceiling(inc1) 
    ## Date of report
    d1   <- o1 + sample(0:input$maxdelay, size = sum(n1), prob = p, replace = TRUE)
    ## 
    
    gen_i_c <- rep(gen_i, length(i1))
    dat1 <- cbind(i1, o1, d1, gen_i_c)
    ###
    dat  <- rbind(dat, dat1)
  }
  
  dat
  
})

renderPlot({
  
  ## summary plot
  # Curves
  as_tibble(dat_all()) %>%
    rename("Daily Infections" = i1, 
           "Daily Onsets" = o1, 
           "Daily Reports" = d1) %>%
    pivot_longer(cols = c("Daily Infections", "Daily Onsets", 
                          "Daily Reports")) %>%
    group_by(name, gen_i_c, value) %>% tally() %>%
    mutate(is_gen_1 = gen_i_c == 1) %>%
    ggplot(.) +
    geom_bar(aes(x = value, fill = gen_i_c, y = n, color = is_gen_1, lwd = is_gen_1), 
             position = 'stack', stat = 'identity') + 
    facet_wrap(~name, nrow = 1) + 
    ylab('N') + xlab('Day') +
    scale_fill_continuous(name = 'Gen. #') +
    scale_color_manual(name = "", values = c("transparent", "red")) +
    scale_linewidth_manual(name = "", values = c(0.1, 1)) +
    # scale_fill_brewer(name = 'Gen.#', type = 'seq', 
    #                   direction = -1, palette = 3) + 
    theme_gray() + coord_cartesian(xlim = c(0, input$ld)) +
    theme(legend.position = 'none',
                axis.text = element_text(size = 10),
          axis.title = element_text(size = 14),
          strip.text = element_text(size = 14))
  
}, height = 200 )
```

Now you have **line-list** data for every person within the modeling window.

